---
title: "Duckweeds and microbes:  \na new frontier in symbiosis research"
date: "Tentative meeting program as of `r format(Sys.time(), '%b %d, %Y')`"
output: github_document
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, dev='png'}
library(here)
htmltools::img(src = knitr::image_uri(here("Slide1.jpg")), alt = 'logo')
```

## Meeting Location

The meeting will take place in the Upper Library at [Massey College](https://www.masseycollege.ca/), located at [4 Devonshire Place, Toronto ON M5S 2E1](https://goo.gl/maps/Meh8ex9ic5cpZmr6A)

## Monday, December 12

**9:00am Continental Breakfast**

**9:30am Megan Frederickson | University of Toronto**

*Opening remarks: using a tiny symbiosis to answer big questions about how host-microbe interactions evolve*

Lenski once wrote that “evolution is simply the conjunction of ecology and genetics.” It sounds simple enough, but the high dimensionality of genomic and environmental variation generates staggering combinatorial complexity. Symbiosis researchers often want to measure traits, fitness, or allelic turnover across many combinations of host genomes, symbiont (meta-)genomes, and environments, but we are typically limited to small-scale experiments in available model systems. My lab members and collaborators, many of whom are also speaking at this meeting, have developed a new experimental system to manipulate many host genotypes and microbiomes across environmental gradients at high throughput using tiny, fast-growing aquatic plants (duckweeds) and their microbiota. I will describe the results of several recent experiments in this system that quantify the main and interactive effects of host genetic variation, microbial community composition variation, and environmental variation for host and microbe fitness, thereby shedding light on how host-microbe interactions evolve. 

**9:55am Anna O'Brien | University of New Hampshire**

*Host plasticity and the microbiome underlie acquired salt tolerance in a freshwater plant*

When faced with rapid environmental changes, species will have to respond rapidly to persist. In addition to evolutionary genetic change, a plethora of plastic mechanisms can underlie species responses, and can occur within a single individual, or across generations. Microbiomes can be directly affected by environmental stressors or indirectly via effects on hosts, and ecological or evolutionary change in microbiomes might contribute to environmental responses of their hosts. One major source of novel stress is the expansion of the urban environment. Urban aquatic organisms face increasingly salty conditions, due to high road salt use. Duckweed (Lemna minor) is a tiny floating freshwater plant that persists across extreme rural to urban ecotones, including salty ponds. We manipulated the presence and absence of the microbiome, and interfered with DNA cytosine methylation (a mechanism of transgenerational plasticity) to quantify the contribution of each to duckweed persistence in salty conditions. Lines of duckweed from more urban sites tolerated higher levels of salt than duckweeds collected from rural areas with low road densities. However, when we removed microbiomes or interfered with DNA cytosine methylation, salt tolerance among lines was similar. Acute salt stress can plastically increase the tolerance of even relatively salt tolerant urban duckweeds over only a few generations in clonal lines. This acute salt-tolerance trades off with growth in salt-free conditions, and occurred regardless of microbiome presence or interference with normal DNA cytosine methylation.

**10:20am Todd Michael | The Salk Institute for Biological Studies**

*The Lemna super pangenome*

As some of the smallest and fastest growing plants on Earth, duckweeds typically grow through the asexual budding, however some species readily flower, leading to the opportunity to find both auto- and allo-polyploids. Also known as the common duckweed, Lemna minor has world-wide distribution and genome sizes that span several hundred megabases (Mb). While Lemna has several chromosomal counts (2n=20,40,42,50,63,126), its ploidy remains poorly characterized. We sequenced 54 L. minor clones from the original duckweed collection and found five classes of genomes representing diploid, triploid, and tetraploids using k-mer analysis, genome assembly and chromosome counting. A reference-free pangenome showed that the Lemna clones represented several hybrids between Lemna minor and Lemna turionifera. In addition, the super pangenome revealed the existence of a new species closely related to Lemna minor but with geographical restriction.

**10:45am Break**

**11:10am Blair Subbaraman | University of Washington**

*Custom workflow automation using Jubilee: duckweed growth assays as a case study*

Machine automation can boost precision, reproducibility and efficiency of research workflows. Open source tools could allow researcher scientists to automate even niche workflows at relatively low cost. Jubilee is an existing open source tool-changing multi-motion platform that we previously developed to support digital fabrication workflows. We will discuss the application of Jubilee to automate duckweed growth assays. We developed novel toolheads, a set of Jupyter notebooks, and associated libraries that can be used to design the experiment and then prompt the machine to set up the sample plates, gather data and then finally to execute relevant data analysis. Jubilee has an active user-developer community including a limited number of research scientists. We discuss the limitations and opportunities for development with this platform as well as providing directions on how to use or develop the Duckbot or other Jubilee-based systems.

**11:35am Jessie Wang | University of Toronto**

*Testing the outcomes of host-microbiome interactions across 1000 environments*

As human activities continue to change the environment at an unprecedented rate, organisms will face novel challenges in their surroundings. The excess deposition of nutrients that often result from human activities may affect the outcomes of species interactions. At low concentrations, these compounds may have positive effects on plant growth and performance. However, at higher concentrations, nutrient loading can impose stress and hamper plant performance. In this study, we tested the outcomes of duckweed-microbiome interactions across environmental gradients of nitrogen, phosphorus, and potassium. This allowed a fine-tuned resolution of the variation in plant responses to environmental stressors, as well as the identification of “optimum” nutrient levels for plant performance. In addition, we investigated the impact of nutrient stress on the composition and diversity of the duckweed microbiome.

**12:00pm Lunch**

**1:30pm Melissa Chen | University of British Columbia**

*Engineered systems in microbial ecology*

In recent years, microbiome research has gained traction and popularity across diverse host systems. Thus, there is an increasing need to develop custom and diverse microbial tools to ask questions about the ecology of host-microbe relationships. In this talk, I will share two examples of engineered microbial systems I have developed in my PhD and Postdoc. In my PhD, I designed and built a biofilm microcosm that features porous biofilm scaffolding through which nutrients diffuse, thus mimicking the nutrient-rich surface of host tissue. I use this microcosm, and isolates from the amphibian system, to show that bacterial biofilms deter fungal pathogen establishment through both space pre-emption and anti-fungal compound production. I also present preliminary data from my Postdoc that showcases a high-throughput plant growth system I am developing. Custom micro-planters (created through 3D printing) are used to grow 'Arabidopsis thaliana' in 96-well format. The 96-well format is compatible with liquid handling robots, which can mix and dilute isolates with high accurancy and allows for rapid screening of pathogenic or beneficial effects by bacteria on plants. A pilot version of this system is being used to screen wild rhizosphere isolates for pathogenic and protective effects on 'A. thaliana'.

**1:55pm Na Wei | Holden Arboretum**

*Plant microbiomes: from terrestrial to aquatic systems*

Understanding the principles of microbiome assembly and function is critical for harnessing microbiota for organismal resilience under changing environments. Our previous work on terrestrial plants especially strawberries and wild relatives has revealed that: (1) host species and genotypes play important roles in the assembly of bacterial and fungal microbiomes both belowground and aboveground; (2) biotic interactions between plants and pollinators influence microbial diversity and network in the anthosphere; (3) microbiome functions are context dependent, such as benefiting plants under drought and harming plants under well-watered conditions; (4) hosts with different ploidy levels differ in their abilities to harness microbiota for stress tolerance. Compared to most terrestrial plants, duckweeds have unique advantages for studying the rules that govern plant–microbiome symbiosis, due to extremely short generations, tiny sizes, and ease in microbiome manipulation. Using Spirodela and Lemna, we are exploring how the evolutionary histories of plants and microbiomes influence symbiosis, and whether warm-adapted microbiomes can help plants respond to a warming climate. 

**2:20pm Eric Lam | Rutgers, the State University of New Jersey**

*Approaches to optimize the detection and quantification of bacteria-duckweed associations*

To study the interaction between duckweed and various microbes, methods to reliably detect and quantify specific microbial strains of interest in the context of the plant host would be invaluable.  Over the past 5 years, we have systematically optimized several techniques to track duckweed associated bacteria strains at varying degrees of specificity with increasing precision.  In this talk, I will summarize our effort in DNA isolation, microbial detection using generic or strain-specific PCR primer sets, and a qualitative attachment assay to compare the mode of interaction between various bacteria isolates with the duckweed Lemna minor.

**2:45pm Erin Baggs | University of California Berkeley**

*Good and evil in the UC Berkeley botanical garden: one Pseudomonad protects duckweed from a rival*

Duckweeds are successful in colonizing diverse environments, despite their lack of conventional immune pathways. It is unclear how duckweeds thrive in the absence of these immune pathways. We investigated the effect of bacteria from duckweeds’ natural habitat on disease progression utilizing the duckweed-Pseudomonas pathosystem.

**3:10pm Break**

**3:40pm Discussion**

**5pm End**

## Tuesday, December 13

**9:00am Continental Breakfast**

**9:30am Jason Laurich | University of Toronto**

*Microbiomes are more and less than the sum of their parts*
 
Microbiomes can benefit hosts by conferring resistance to pathogens, improving stress tolerance, or promoting growth. We explored host-microbiome interactions using the common duckweed Lemna minor and a simplified synthetic microbiome. We evaluated the effects of individual microbial strains and entire microbial communities on duckweed growth and microbial cell density. Across three populations of L. minor, we observed substantial variation in strain effects on host growth and found that the effects of synthetic microbial communities could not be predicted from individual strains. In most cases, simplified microbial communities were more productive, and more beneficial to plant hosts, than the mean (additive) effect of individual strains. In one population, we instead found that the presence of a single highly pathogenic strain likely determined the sub-additive effect of its microbiome. These results highlight the importance of evaluating the effects of microbiomes on host plants at both the individual strain level and at the level of whole communities, where cooperative, non-additive internal dynamics may shape their effects on host plants. We also evaluated the extent of fitness alignment between microbes and duckweeds and found evidence of broadly correlated fitness interests. However, the strength of this correlation varied across microbes, and the fitness of some of the most beneficial bacteria were neutrally, or even negatively correlated with L. minor fitness. Our results speak to the complexity of community dynamics in microbiomes, and the extent to which interactions within them can alter both the selective landscape of host-microbiome interactions, and their effects on host plants.

**9:55am Amy Angert | Departments of Botany and Zoology, UBC**

*Duckweed mesocosms for range dynamics*

Abstract forthcoming. 

**10:20am Martin Turcotte | University of Pittsburgh**

*The impact of duckweed polyploidy on species interactions*

Whole-genome duplication is a common macromutation across the tree of life and especially common in plants. It can cause diverse and sometimes extensive phenotypic changes but the impact of polyploidy on other community members has received less attention. Using synthetic autopolyploid duckweed, we found that polyploidy alters microbiome richness, diversity, and the composition of the ‘core’ microbiome. We also report on another project testing how duckweed polyploids differ in herbivore tolerance and sometimes influences herbivore performance.  

**10:45am Break**

**11:10am Discussion**

**12:00pm Lunch**

**2:00pm Mark Jewell | McGill University**

*Phenotypic and fitness consequences of microbial associations for Lemna minor*

Much recent work has focused on characterising the mutualistic elements of the duckweed microbiome, often aiming to identify bacterial strains that can increase plant fitness. Here we assess the fitness and phenotypic consequences of the full intact microbiome for L. minor by assaying plants from several sites, with and without their natural microbiomes, over a range of environmental conditions. We find that the overall effect of the microbiome is to suppress plant fitness which is accompanied by several phenotypic modifications.

**2:25pm Takuji Usui | Biodiversity Research Centre, University of British Columbia**

*The evolution of coexistence across the species boundary*

The entirety of Earth’s ~8.7 million extant species are the outcome of two processes: the creation of new lineages by speciation, and the persistence of these lineages by ecological coexistence. To understand the origins of biodiversity, evolutionary biology has primarily focused on the causes and means of speciation. However, we know less about the evolution and mechanisms of coexistence across diverging lineages. Here, we quantify the evolutionary trajectory of coexistence within diverging lineages, to across the species boundary. We parameterized coexistence by quantifying the mutual invasion criteria in ~3000 invasion experiments across 127 unique lineages of the duckweeds Spirodela polyrhiza and Spirodela intermedia. We present how coexistence mechanisms (i.e., differences in niche and competitive ability) evolve with genetic and spatial distance. By considering how persistence evolves across diverging lineages, we begin to unravel how coexistence mechanisms themselves can operate as a route for speciation and the origin of coexisting species.

**2:50pm Jiaqi Tan | Department of Biological Sciences, Louisiana State University**

*Trophic interactions in microbiomes determine plant host fitness*

Plant microbiomes are composed of various microorganisms, including prokaryotes, eukaryotes, and viruses. They are the key determinant of plant host fitness. However, little is known about how interactions, especially trophic interactions, between microbes from different domains modify the importance of microbiomes for plant hosts. Using duckweed Lemna minor and its microbiomes, we experimentally examined the influences of host-parasitism and predator-prey interactions within the microbiomes on the multi-generational fitness of L. minor. When compared to the axenic control, bacterial inoculation did not affect L. minor fitness. However, the presence of parasites reduced host fitness whereas the presence of predators increased host fitness. These results highlight the importance of knowledge of microbial trophic interactions for predicting the outcomes of plant-microbiome interactions.

**3:15pm Break**

**3:45pm Discussion**

**4:45pm Closing remarks**
